Phylogenetics and the human microbiome.

### 2014-04-23 sbreview.tex
Fixen sie formatten for arxiv.
Edits from read-through.

The human microbiome is the ensemble of genes in the microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiota, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiota}
The human microbiota is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiota have remarkable metabolic potential, being an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the disruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
In this section I will briefly review what is known about the human microbiota and its effect on our health.

@@

In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys.''
This terminology is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.
Such surveys would ideally return a census of all of the microbes in a sample along with their abundances.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy, and several groups have built databases of core gene families and provide programs and/or web tools to subset a metagenome to that core set of genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using one of the methods described below.

Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.

\subsection{Whole genomes}
Whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
This data may become common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.

@@

Phylogenetic inference has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
The clearest such example is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree which is annotated as follows.
The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades.
where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

It is not straightforward to define a clear notion of optimality for OTU inference.
In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
Internal measures evaluate the clustering in various ways without reference to a ``true'' clustering, with the general idea that within-cluster distances should be small compared to between-cluster distances.

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.

The centrality of the OTU concept can be seen by the fact that the by-sample table of OTU observations (i.e.\ the matrix of counts with rows representing samples and columns representing OTUs) is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
With the exception of the work of \citet{sharpton2011phylotu} and \citet{Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we describe their application.

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'' required to move piles of dirt (each dirt pile configuration sized in proportion to one probability distribution) in one configuration to another along the tree, where work is defined as mass times distance.
In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.

>>>>

The human microbiome is the ensemble of genes in the microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing and genetic analysis of microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiota, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiota}
The human microbiota is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiota have remarkable metabolic potential, being an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}.
Our food and drug intake, in turn, impacts the diversity of microbes present.
Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding \citep[reviewed in][]{funkhouser2013mom}.
In this section I will briefly review what is known about the human microbiota and its effect on our health.

@@

In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys.''
This terminology is equivalent to the ``barcoding'' terminology more commonly used for eukaryotic surveys using 18S or the fungal internal transcribed spacer (ITS).
Such surveys would ideally return a census of all of the microbes in a sample along with their abundances.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
This can distort inferences concerning actual microbe abundances based on read copy number.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts method \citep{felsenstein1985phylogenies} to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to the organism from which it came.
Because metagenomes do not proceed through an amplification step, they do not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.

It is possible to subset metagenomic data to marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy, and several groups have built databases of core gene families as well as provided programs and/or web tools to phylogenetically analyze metagenomes subset to those core genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using one of the methods described below.

Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is the sequencing of mRNA in bulk which is called ``metatranscriptomics".

\subsection{Whole genomes}
In addition to conventional genome sequencing for microbial genomes, whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
This data may become more common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.

@@

Phylogenetic inference has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
The clearest such example is the discovery that archaea, although similar to bacteria in their gross morphology, form their own separate lineage \citep{woese1977phylogenetic}.

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy \citep{kreig1984bergey}, although the actual revision process appears opaque.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree, which is annotated as follows.
The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades, where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as an allowance for either phylogenetic error, lack of resolution of the 16S gene, or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI or EMBL sequence databases.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the taxonomic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (again, for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

It is not straightforward to define a clear notion of optimality for OTU inference.
While in phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on an inferred clustering.
External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
Internal measures evaluate the clustering in various ways without reference to a ``true'' clustering, with the general idea that within-cluster distances should be small compared to between-cluster distances.

There are many OTU inference methods with various speeds and strategies.
Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
\citet{Wang2013-ol} optimize a criterion of cluster modularity.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.

The centrality of the OTU concept can be seen by the fact that the by-sample table of OTU observations (i.e.\ the matrix of counts with rows representing samples and columns representing OTUs) is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
With the exception of the work of \citet{sharpton2011phylotu} and \citet{Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies of the OTUs.
The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used.
However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.

@@

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a sublinear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    Assume that the sequence data to build the phylogenetic tree derives from two samples: the green sample and the blue sample.
    When the samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch length that sits ancestral to clades that are \textit{uniquely} composed of one sample or another, compared to when they are separate (right tree).
    The bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample (green or blue, but not brown) divided by the total branch length.
    The ratio is smaller when the samples are interspersed (left) than they are when separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[32]{r}{2.2in}
  % \capstart
  \vspace{-26pt}
  \begin{center}
    \includegraphics[width=2.2in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}
